---
title: "For loop practice"
author: "Abby Lewis"
date: "2023-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Several exercises for practicing for loops, primarily drawn from [R-bloggers](https://www.r-bloggers.com/2018/03/loops-in-r-exercises/)

## Exercise 1

Write a for loop that iterates over the numbers 1 to 7 and stores the cube of each number in `output`.

```{r}
#output <- ??? #how long do you want the output to be? Can you create a vector of that length?
#for () { #what do you want to iterate over?
  
  #BODY

#} end for loop
#output #view output
```

seven <- c(1,2,3,4,5,6,7)
for(i in 1:length(seven)) {}




## Exercise 2

Write a for loop that iterates over the column names of the inbuilt iris dataset and print each together with the number of characters in the column name in parenthesis. Example output: "Sepal.Length (12)". 
Use the following functions print(), paste0() and nchar().

```{r}
iris
?paste0()
?nchar()
```

## Exercise 3

Using a for loop simulate the flip a coin twenty times, keeping track of the individual outcomes (1 = heads, 0 = tails) in a vector that you preallocate.

```{r}
sample(c("heads","tails"),1)
```

## Exercise 4

Use a nested for loop (a for loop inside a for loop) to produce the following matrix. Preallocate the matrix with NA values.
    0     1     2     3     4
    1     0     1     2     3
    2     1     0     1     2
    3     2     1     0     1
    4     3     2     1     0
    
```{r}
output <- matrix(NA, nrow = 5, ncol = 5) #Start with this matrix
# For loop here
```

## Exercise 5

The first two terms of the Fibonacci sequence are both ones. Subsequent terms of the sequence are found by summing the two terms immediately previous. Write a for loop that produces the first 10 terms of the Fibonacci sequence

```{r}

```

