---
title: "Functions_2"
author: "Abby Lewis"
date: "2023-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises

### Beginner

1. What are the three important components of a function?

Formals are the list of arguments that control how you call the function

The body is the code within the function

The environment is the data structure that determines how the function finds the values associated with the name

2. Write a function to calculate the z-score of a vector, where the z-score is 
(value - mean)/sd

```{r}
z <- c(10, 20, 30, 40, 50)
mean_z <- mean(z)
sd_z <- sd(z)

#' Title
#'
#' @param z c(10, 20, 30, 40, 50)
#'
#' @return z scores for vector "z"
#' @export
#'
#' @examples see below

score <- function(z) {
  (z-mean_z)/sd_z
}

score(10)
```

3. Add documentation to your function in the `roxygen2` format

### Intermediate

`map(1:3, ~ runif(2))` is a useful pattern for generating random numbers, but `map(1:3, runif(2))` is not. Why not? Can you explain why it returns the result that it does?

```{r}
map(1:3, ~ runif(2))
map(1:3, runif(2))
```

2. Create infix versions of the set functions `intersect()`, `union()`, and `setdiff()`. You might call them `%n%`, `%u%`, and `%/%` to match conventions from mathematics.

```{r}

```


### Advanced

1.  Use `map()` to fit linear models to the `mtcars` dataset using the formulas stored in this list:

```{r}
formulas <- list(
  mpg ~ disp,
  mpg ~ I(1 / disp),
  mpg ~ disp + wt,
  mpg ~ I(1 / disp) + wt
)
```

2. Rewrite the following code snippets into prefix form:

```{r, eval = FALSE}
1 + 2 + 3

1 + (2 * 3)

if (length(x) <= 5) x[[5]] else x[[n]]
```